# Reproducible Research: Peer Assessment 1
### R Markdown File by PCommodore

The data for this assignment can be downloaded from the course web site:
Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken
* The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Step 1: Loading and preprocessing the data

This step consists of following:
* 1.1 Read in the file as a data frame
* 1.2 Convert date column into column of the Date class
* 1.3 Remove all rows with NA

```{r echo=TRUE}
#1.1
activity=read.csv("C:/Users/LenovoJ/Desktop/repos/RepData_PeerAssessment1/activity/activity.csv")
#1.2
activity$date=as.Date(activity$date,"%Y-%m-%d")
#1.3
activitycleaned=na.omit(activity)
```

```{r echo=TRUE}
head(activity)
```

## Step 2: What is mean total number of steps taken per day?

This step consists of the following:
* 2.1 Compute data frame presenting total number of steps by day
* 2.2 Plot histogram of total number of steps across day
* 2.3 Compute data frame presenting mean number of steps by day
* 2.4 Compute data frame presenting median number of steps by day

```{r echo=TRUE}
library(reshape2)
library(lattice)
#2.1
activitymelt=melt(activitycleaned,id=c("date"),measure.vars=c("steps"))
activitydcastsum=dcast(activitymelt,date~variable,sum)

#2.2
plot(activitydcastsum$date,activitydcastsum$steps,type='h',xlab="Day",ylab="Sum of Steps",main="Histogram for Sum of Steps by Day")

#2.3
activitydcastmean=dcast(activitymelt,date~variable,mean)
ameanbyday=activitydcastmean

#2.4
activitymelt$value=as.double(activitymelt$value)
activitydcastmedian=dcast(activitymelt,date~variable,median)
amedianbyday=activitydcastmedian
```

```{r echo=TRUE}
#Data frame for total number of steps by day
head(activitydcastsum)
#Data frame for mean number of steps by day
head(ameanbyday)
#Data frame for median number of steps by day
head(amedianbyday)

```



## Step 3: What is the average daily activity pattern?

This step consists of the following:
* 3.1 Time Series plot of the 5-minute interval (x-axis) and the average number of steps, averaged across all days over (y-axis)
* 3.2 State the interval which contains the maximum number of steps

```{r echo=TRUE}
#3.1
activitymeltinterval=melt(activitycleaned,id=c("interval"),measure.vars=c("steps"))
activitydcastintervalmean=dcast(activitymeltinterval,interval~variable,mean)
ameanbyinterval=activitydcastintervalmean

#Plot Time Series 
plot(ameanbyinterval$interval,ameanbyinterval$steps,type='l',xlab='Interval',ylab='Average Number of Steps',main='Time Series Plot Mean Steps across Intervals')

#3.2
#The maximum number of steps occur at interval 0835-0840
ameanbyinterval[ameanbyinterval$steps==max(ameanbyinterval$steps),]
```


## Step 4: Imputing missing values

This step consists of the following:
* 4.1 Find the number of rows with NA
* 4.2 Fill in missing values with the mean number of steps for the respective intervals
* 4.3 Compute new data frame with the missing values filled in
* 4.4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r echo=TRUE}
#4.1
numbermissing=nrow(activity)-nrow(activitycleaned)
numbermissing

#4.2
for (i in 1:nrow(activity)) {
if (is.na(activity$steps[i])){
activity$filledsteps[i]=ameanbyinterval[ameanbyinterval$interval==activity$interval[i],][1,2]}
else {activity$filledsteps[i]=activity$steps[i]}
}


#4.3
newdata=activity
newdata$steps=newdata$filledsteps
newdata=newdata[,1:3]


#4.4
newdatamelt=melt(newdata,id=c("date"),measure.vars=c("steps"))
newdatadcastsum=dcast(newdatamelt,date~variable,sum)

#Plot histogram for Sum of Steps per day
plot(newdatadcastsum$date,newdatadcastsum$steps,type='h',xlab="Day",ylab="Sum of Steps",main="Histogram for Sum of Steps by Day (Missing Values Filled)")

#Data frame for mMean steps per day
newdatadcastmean=dcast(newdatamelt,date~variable,mean)
newmeanbyday=newdatadcastmean

#Data frame for median steps per day
newdatamelt$value=as.double(newdatamelt$value)
newdatadcastmedian=dcast(newdatamelt,date~variable,median)
newmedianbyday=newdatadcastmedian
```

## Step 5: Are there differences in activity patterns between weekdays and weekends?

This step consists of the following:

* 5.1 Create new factor 'weekday' and 'weekend'
* 5.2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r echo=TRUE}
#5.1
newdata$day=weekdays(newdata$date)
newdata$weekday=newdata$day

for (i in 1:nrow(newdata)) {
if (newdata$day[i]=="Saturday"|newdata$day[i]=="Sunday"){
newdata$weekday[i]="weekend"}
else {newdata$weekday[i]="weekday"}
}

weekdaydata=newdata[newdata$weekday=="weekday",]
weekenddata=newdata[newdata$weekday=="weekend",]

weekdaymeltinterval=melt(weekdaydata,id=c("interval"),measure.vars=c("steps"))
weekdaydcastintervalmean=dcast(weekdaymeltinterval,interval~variable,mean)
weekdaymeanbyinterval=weekdaydcastintervalmean
#plot(weekdaymeanbyinterval$interval,weekdaymeanbyinterval$steps,type='l')

weekendmeltinterval=melt(weekenddata,id=c("interval"),measure.vars=c("steps"))
weekenddcastintervalmean=dcast(weekendmeltinterval,interval~variable,mean)
weekendmeanbyinterval=weekenddcastintervalmean
#plot(weekendmeanbyinterval$interval,weekendmeanbyinterval$steps,type='l')

weekdaylabel=rep("weekday",nrow(weekdaymeanbyinterval))
weekdaymeanbyinterval$label=weekdaylabel

weekendlabel=rep("weekend",nrow(weekendmeanbyinterval))
weekendmeanbyinterval$label=weekendlabel

weekdaynweekend=rbind(weekdaymeanbyinterval,weekendmeanbyinterval)

#5.2
attach(weekdaynweekend)
xyplot(steps~interval|label,main="Time Series Panel Plot by Weekday/Weekend",xlab="Interval",ylab="Steps",type='l',layout=c(1,2))
```

### End of File